---
title: "Interactive map tutorial"
author: "Jahred Liddie"
date: "2023-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(leaflet) # this the the interactive map package we'll use
library(sf)
library(ggmap) # good for basemaps, also allows access to Google Maps API

# read practice datasets
libraries <- st_read("https://bostonopendata-boston.opendata.arcgis.com/datasets/boston::public-libraries.geojson?where=1=1&outSR=%7B%22latestWkid%22%3A2249%2C%22wkid%22%3A102686%7D")

foodtrucks <- st_read("https://bostonopendata-boston.opendata.arcgis.com/datasets/boston::food-truck-schedule.geojson?where=1=1&outSR=%7B%22latestWkid%22%3A3857%2C%22wkid%22%3A102100%7D")

nhoods <- st_read("https://bostonopendata-boston.opendata.arcgis.com/datasets/boston::boston-neighborhood-boundaries-approximated-by-2020-census-block-groups.geojson?where=1=1&outSR=%7B%22latestWkid%22%3A2249%2C%22wkid%22%3A102686%7D")

```

# Task 1: simple map

First, check out what the `sf` dataframes look like compared to typical dataframes. Then, we'll review how to make a basic, static map in R using `ggplot`.

```{r staticmap1}

# basic map
ggplot(nhoods) +
  geom_sf(alpha = 0.75, linewidth = 0.5, color = "black") +
  geom_sf(data = libraries, color = "blue") +
  labs(title = "Libraries in Boston") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```

You can also quickly add a basemap using `ggmap`.

```{r staticmap2}
# it's also possible to search the API for Boston's coordinates, 
  # but here they are:
boston_basemap <- get_stadiamap(c(left = -71.2, bottom = 42.22, 
                                  right = -70.95, top = 42.4), zoom = 12)

# basic map
ggmap(boston_basemap) +
  geom_sf(data = nhoods, alpha = 0.75, 
          linewidth = 0.5, color = "black", inherit.aes = FALSE) +
  geom_sf(data = libraries, color = "blue", inherit.aes = FALSE) +
  labs(title = "Libraries in Boston") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

# Task 2: A simple interactive map using `leaflet`

There are other methods to create interactive maps in R, but we'll start with a simple approach using the `leaflet` package.

In the plot below, we plot the Boston neighborhood dataset again with locations of the libraries on top of them. We allow the Boston neighborhoods to have a "highlight" color when our cursor hovers over them. We also put a basemap behind our plotted data.

```{r intmap1}

# in R, leaflet generally has gpplot's syntax / structure, 
  # but it can use pipes (%>%)
leaflet(nhoods) %>% # this is similar to 'ggplot' 
  
  # this is similar to 'geom_sf'; weight controls line thickness
  # you can also remove borders
  addPolygons(group = "Boston neighborhoods", weight = 2, color = "black", 
              fillColor = "blue", fillOpacity = 0.1,
              highlightOptions = highlightOptions(fillColor = "darkblue", 
                                                  fillOpacity = 1)) %>%
  
  # this is more specific to points - note the different arguments
  addCircles(data = libraries, group = "Libraries",
             color = "red", opacity = 1, radius = 0.5) %>%
  
  # this is similar to ggmap; see `providers' for a list of options
  addProviderTiles(providers$CartoDB.Positron)

```

# Task 3: A more complicated interactive map using `leaflet`

In the plot below, we extend our previous plot by adding labels for polygons and text popups for points. We limit the plot view for our audience. We also include control options, such that certain layers (which we specify) can be added and removed from the map.

```{r intmap2}

# easy way to retrieve bounds of an 'sf' object
limits <- st_bbox(nhoods)

leaflet(nhoods) %>% 
  
  addPolygons(group = "Boston neighborhoods", weight = 2, color = "black", 
              fillColor = "blue", fillOpacity = 0.1,
              label = nhoods$blockgr2020_ctr_neighb_name,
              highlightOptions = highlightOptions(fillColor = "darkblue", 
                                                  fillOpacity = 1)) %>%
  
  # now we also add very simple popups
  addCircles(data = libraries, group = "Libraries",
             popup = libraries$BRANCH,
             color = "red", opacity = 1, weight = 5) %>%
  
  addCircles(data = foodtrucks, group = "Food trucks",
             popup = foodtrucks$Truck,
             color = "green", opacity = 1, weight = 4) %>%
  
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # these are the control options for the map
    # note that tiles can also be controlled
  addLayersControl(
    position = "bottomright",
    baseGroups = "Boston neighborhoods", # needs to match `group` name above
    overlayGroups = c("Libraries", "Food trucks")) %>%
  
  # now add bounds to map to limit scrolling
  setMaxBounds(lng1 = as.numeric(limits[1]) - 0.1,
               lat1 = as.numeric(limits[2]) - 0.1,
               lng2 = as.numeric(limits[3]) + 0.1,
               lat2 = as.numeric(limits[4]) + 0.1
              )

```

# More resources

See [here](https://rstudio.github.io/leaflet/) for a more in-depth tutorial of `leaflet`, including how to integrate them with `RShiny`.
